A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2011-06-09 contactpaper/paper.tex
a few more changes to contact-density paper
Tested-on: bingley (Linux 2.6.32-5-amd64)

  First, let's rewrite the free energy in terms of a smaller set of
  weighted densities:
  \begin{align}
    A_{HS} &= \int \left\{ -n_0 \ln\left( 1 - n_3\right)
    + \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
    + \frac{
      \frac13 n_2^3 - n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}
    }{
      8\pi \left( 1 - n_3 \right)^2
    }\right\}
    \\
    &= \int \left\{ -\frac{n_2}{4\pi R^2} \ln\left( 1 - n_3\right)
    + \frac{1}{4\pi R} \frac{n_2^2 - \mathbf{n}_{V2} \cdot\mathbf{n}_{V2}}{1-n_3}
    + \frac{
      \frac13 n_2^3 - n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}
    }{
      8\pi \left( 1 - n_3 \right)^2
    }\right\}
\end{align}

\begin{align}
    \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')} &=
    \frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    + \frac{1}{4\pi}\frac{
      \frac13 n_2^3 - n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}
    }{
      (1-n_3(\mathbf{r}'))^3
    }
    \\
    \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')} &=
    -\frac{1}{4\pi R^2} \ln\left( 1 - n_3\right)
    +  \frac{2n_1}{1-n_3}
    + \frac{n_2^2 - \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}}{8\pi \left( 1 - n_3 \right)^2}
    \\
    \frac{\delta A_{HS}}{\delta \mathbf{n}_{V2}(\mathbf{r}')} &=
    \frac{2\mathbf{n}_{V1}}{1-n_3}
    + \frac{n_2 \mathbf{n}_{V2}}{4\pi \left( 1 - n_3 \right)^2}
    \\
    \frac{\delta A_{HS}}{\delta R(\mathbf{r}')} &=
    \frac{n_2}{2\pi R^3} \ln\left( 1 - n_3\right)
    - \frac1{R} \frac{n_1n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
  \end{align}

@@

So we need to find the small change in free energy that is caused by
these small changes in the shape and size of the hard spheres that are
in contact with this point.  This changes the weighted densities in
the following way---assuming that our point of interest is
$\mathbf{r}$.
\begin{align}
  \Delta n_3(\mathbf{r}') &= n(\mathbf{r})\Delta A \Delta R
  \delta(|\mathbf{r}-\mathbf{r}'| - R)
   \\
  \Delta n_2(\mathbf{r}') &=
  \left(n\left(\mathbf{r} + \Delta R
  \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\right)\left(1+2\frac{\Delta
  R}{R}\right)- n(\mathbf{r})\right) \Delta A
  \delta(|\mathbf{r}-\mathbf{r}'|-R) \\
  &=
  \left(\frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\cdot
    \mathbf{\nabla}n(\mathbf{r}) + 2 \frac{n(\mathbf{r})}{R}
    \right) \Delta A \Delta R
  \delta(|\mathbf{r}-\mathbf{r}'|-R) \\
  \Delta \mathbf{n}_{V2}(\mathbf{r}') &=
  \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}
  \left(n\left(\mathbf{r} + \Delta R
  \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\right)\left(1+2\frac{\Delta
  R}{R}\right)- n(\mathbf{r})\right) \Delta A
  \delta(|\mathbf{r}-\mathbf{r}'|-R) \\
  &=
  \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}
  \left(\frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\cdot
    \mathbf{\nabla}n(\mathbf{r}) + 2 \frac{n(\mathbf{r})}{R}
    \right) \Delta A \Delta R
  \delta(|\mathbf{r}-\mathbf{r}'|-R) 
\end{align}
\end{widetext}

@@

>>>>

  First, let's rewrite the free energy in terms of a smaller set of
  weighted densities:
  \begin{align}
    A_{HS} &= \int \left\{
      -n_0 \ln\left( 1 - n_3\right)
      + \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
      + (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
        n_3 + (1-n_3)^2 \ln(1-n_3)
      }{
        36\pi n_3^2(1-n_3)^2
      }
    \right\}
    \\
    &= \int \left\{
      -\frac{n_2}{4\pi R^2} \ln\left( 1 - n_3\right)
      + \frac{1}{4\pi R} \frac{n_2^2 - \mathbf{n}_{V2} \cdot\mathbf{n}_{V2}}{1-n_3}
      + (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
        n_3 + (1-n_3)^2 \ln(1-n_3)
      }{
        36\pi n_3^2(1-n_3)^2
      }
    \right\}
\end{align}

\begin{align}
    \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')} &=
    -\frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    + \frac{n_2^3 -
      3n_2\mathbf{n}_{V2}\cdot\mathbf{n}_{V2}}{36\pi}\left(
      \frac{1}{(1-n_3)^2} + 2\frac{\ln(1-n_3)}{n_2^2(1-n_3)}
    \right)
    \\
    \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')} &=
    -\frac{1}{4\pi R^2} \ln\left( 1 - n_3\right)
    +  \frac{2n_1}{1-n_3}
    + 3(n_2^2 - \mathbf{n}_{V2}\cdot\mathbf{n}_{V2})\frac{n_3 +
      (1-n_3)^2\ln(1-n_3)}{
      36\pi n_3^2(1-n_3)^2
    }
    \\
    \frac{\delta A_{HS}}{\delta \mathbf{n}_{V2}(\mathbf{r}')} &=
    \frac{2\mathbf{n}_{V1}}{1-n_3}
    - 6 n_2 \mathbf{n}_{V2} \frac{n_3 +
      (1-n_3)^2\ln(1-n_3)}{
      36\pi n_3^2(1-n_3)^2
    }    \\
    \frac{\delta A_{HS}}{\delta R(\mathbf{r}')} &=
    \frac{n_2}{2\pi R^3} \ln\left( 1 - n_3\right)
    - \frac1{R} \frac{n_1n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
  \end{align}

@@

So we need to find the small change in free energy that is caused by
these small changes in the shape and size of the hard spheres that are
in contact with this point.  In the following section, we will work
out the changes in the weighted densities at position $\mathbf{r}'$
when point of contact we are considering is $\mathbf{r}$.

The packing fraction $n_3$ is changed at the point where the sphere is
extended, since a few more spheres will overlap this point.
\begin{align}
  \Delta n_3(\mathbf{r}') &= n(\mathbf{r})\Delta A \Delta R
  \delta(|\mathbf{r}-\mathbf{r}'| - R)
\end{align}
The density $n_0$ is the average density, averaged over a spherical
shell.  As such, shifting the position of a shell just shifts which
location we are averaging over.
\begin{align}
  \Delta n_0(\mathbf{r}') &=
    \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\cdot
      \mathbf{\nabla}n(\mathbf{r}) \Delta A \Delta R
    \delta(|\mathbf{r}-\mathbf{r}'|-R)
\end{align}
The change in $n_2$ comes from two terms: a change in the density value, as
we saw in $n_0$, and a change in the area of the segment.
\begin{align}
  \Delta n_2(\mathbf{r}') &=
  \left(n\left(\mathbf{r} + \Delta R
  \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\right)\left(1+2\frac{\Delta
  R}{R}\right)- n(\mathbf{r})\right) \Delta A
  \delta(|\mathbf{r}-\mathbf{r}'|-R) \\
  &=
  \left(\frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\cdot
    \mathbf{\nabla}n(\mathbf{r}) + 2 \frac{n(\mathbf{r})}{R}
    \right) \Delta A \Delta R
  \delta(|\mathbf{r}-\mathbf{r}'|-R)
\end{align}
The vector density $\mathbf{n}_{V2}$ is similar to the density
$n_2$: both changes in the density value and changes in the surface
area play a role.
\begin{align}
  \Delta \mathbf{n}_{V2}(\mathbf{r}') &=
  \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}
  \left(n\left(\mathbf{r} + \Delta R
  \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\right)\left(1+2\frac{\Delta
  R}{R}\right)- n(\mathbf{r})\right) \Delta A
  \delta(|\mathbf{r}-\mathbf{r}'|-R) \\
  &=
  \frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}
  \left(\frac{\mathbf{r}-\mathbf{r}'}{|\mathbf{r}-\mathbf{r}'|}\cdot
    \mathbf{\nabla}n(\mathbf{r}) + 2 \frac{n(\mathbf{r})}{R}
    \right) \Delta A \Delta R
  \delta(|\mathbf{r}-\mathbf{r}'|-R) 
\end{align}
The density $n_1$ (and similarly $\mathbf{n}_{V1}$) can be roughly
interpreted as the density weighted by the area-averaged mean
curvature~\cite{oversteegen2005general, roth2010review}.  When we
extend the sphere at the contact point, we will introduce regions of
both positive and negative mean curvature.  In principle, we can do
this in such a way that the two portions cancel, and hence we will
assume that
\begin{align}
  \Delta n_1(\mathbf{r}') &= 0 \\
  \Delta \mathbf{n}_{V1}(\mathbf{r}') &= 0
\end{align}
In fact, the negative and positive portions are spatially separated,
so these terms should probably include a derivative term, which I
should work out.  Particularly as it may even happen to cancel a
different derivative term that we then won't have to drop!

We need to then find the change in Helmholtz free energy due to this
change at position $\mathbf{r}$:
\begin{align}
  \Delta A_{HS} &= \int \mathbf{dr}'
  \frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')}\Delta n_3(\mathbf{r}')
  +
  \frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')}\Delta n_2(\mathbf{r}')
  +
  \frac{\delta A_{HS}}{\delta \mathbf{n}_{V2}(\mathbf{r}')} \cdot \Delta \mathbf{n}_{V2}(\mathbf{r}')
\end{align}

\end{widetext}

@@

\begin{widetext}
  \appendix

  \section{Partial derivatives of the White Bear free energy functional}

  Here is the White Bear free energy functional for the hard-sphere
  fluid:~\cite{roth2002whitebear}
  \begin{align}
    A_{HS} &= \int \left\{
    -n_0 \ln\left( 1 - n_3\right)
    + \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
    + (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
      n_3 + (1-n_3)^2 \ln(1-n_3)
    }{
      36\pi n_3^2(1-n_3)^2
    }
    \right\}
\end{align}

\begin{align}
    \beta\frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')} &=
    -\frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    + \frac{n_2^3 -
      3n_2\mathbf{n}_{V2}\cdot\mathbf{n}_{V2}}{36\pi}\left(
      \frac{1}{(1-n_3)^2} + 2\frac{\ln(1-n_3)}{n_2^2(1-n_3)}
    \right)
\end{align}
\end{widetext}

\begin{align}
    \beta\frac{\delta A_{HS}}{\delta n_0(\mathbf{r}')} &= -\ln(1-n_3)
    \\
    \beta\frac{\delta A_{HS}}{\delta n_1(\mathbf{r}')} &= \frac{n_2}{1-n_3}
    \\
    \beta\frac{\delta A_{HS}}{\delta n_2(\mathbf{r}')} &=
      \frac{n_1}{1-n_3}
      + (n_2^2 - \mathbf{n}_{V2}\cdot\mathbf{n}_{V2})\frac{n_3 +
        (1-n_3)^2\ln(1-n_3)}{
        12\pi n_3^2(1-n_3)^2
      }
\end{align}
\begin{align}
    \beta\frac{\delta A_{HS}}{\delta \mathbf{n}_{V1}(\mathbf{r}')} &=
      \frac{\mathbf{n}_{V2}}{1-n_3}
    \\
    \beta\frac{\delta A_{HS}}{\delta \mathbf{n}_{V2}(\mathbf{r}')} &=
      \frac{\mathbf{n}_{V1}}{1-n_3}
      - 6 n_2 \mathbf{n}_{V2} \frac{n_3 +
        (1-n_3)^2\ln(1-n_3)}{
        36\pi n_3^2(1-n_3)^2
      }
  \end{align}
